Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-13 paper_jasa.tex
some typo fixed

\begin{table*}[hbtp]
\caption{Comparison of visual inference with conventional inference.}
\centering 
\begin{tabular}{llll} 
\hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\begin{tabular}{rrrrrrrrrrrrrrrrrrr}
$K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$  & $x$ & $p$-value && $K$  & $x$ & $p$-value\\ \cline{1-3} \cline{5-7}\cline{9-11}\cline{13-15}\cline{17-19}
1 &  1 & 0.0500 && 2 &  1 & 0.0975 && 3 & 1 & 0.1426 && 4 & 1 & 0.1855 && 5 & 1 & 0.2262 \\%\cline{1-3}
&&&& 2 &  2 & 0.0025 && 3 & 2 & 0.0073 && 4 & 2 & 0.0140 && 5 & 2 & 0.0226 \\
&&&& &&&& 3 & 3 & 0.0001 && 4 & 3 & 0.0005 && 5 & 3 & 0.0012 \\%\cline{1-6}
&&&&  &    &        &   &   &        &&& 4 & 4 & $< 0.0001$ && 5 & 4 & $< 0.0001$ 
\end{tabular}
\end{center}
\label{pvalue}
\end{table}

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of size $\alpha$, is defined to be 
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot in the lineup, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}) < \alpha$. 
\end{dfn}

@@

\hh{If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p+i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition, see \citet{butler93} for details. Estimating $p_i$ will be one of the main concerns for the remainder of the paper.} One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

@@

Under the null hypothesis $X \sim Binom_{K, 1/m}$, leading to a type I error of:
\[
\text{Power}_V(\theta, K)= 1 - Binom_{K, 11/m} (x_\alpha - 1),
\]
where $x_\alpha$ is defined as in (\ref{dfn:test}), as  $P(X \ge x_{\alpha}) < \alpha$.

@@

\green{Notation changes $i$ to $j$, and $p$ to $\pi$. Can someone please decide which to use, and make the appropriate changes? How does the ``Let $X_{j\ell} \sim B_{1, \pi_{j\ell}}$...'' below relate to the ``let $X_i$ be a binary random variable...'' above? It looks like this is a repeat. One needs to go once notation is clarified.}

\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate actual data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

@@

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for each of the simulation experiments.} % title name of the table
\centering
% centering table
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

% Entering 2rd row
&  5 & 0, 1, 2, 3, 5  & 0.1, 0.4, 0.7, 1, 1.5&\\[-1ex]
\raisebox{1.5ex}{300} & 12
& 1, 3, 5, 7, 10  & 0, 0.8, 1.75, 2.3, 3.5 &\\[1ex]
% [1ex] adds vertical space
\hline
% inserts single-line
\end{tabular}
\label{tbl:experiment_params}
\end{table}

@@

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }%\blue{Notice that we do not have data on one lineup (so we have only 29 lineups) for experiment 3 due to the screening process.}}
\begin{center}
\begin{tabular}{ccrrcc}
   \hline 
 Experiment & Subject & Male & Female & Responses &Lineup\\ 
    \hline
1 & 239 & 121 & 107 & 2249 &  60 \\ 
  2 & 351 & 185 & 164 & 3636 &  70 \\ 
  3 & 155 & 103 &  52 & 1511 &  29 \\  
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

\begin{table}[hbtp]
\caption{Parameter estimates of model in Equation \ref{eqn:mixed}. Estimates are highly significant with $p$-value $<$  0.0001 for all three experiment data.}
\begin{center}
\begin{tabular}{cr@{.}lcc}
  \hline
 &  \multicolumn{3}{c} {Fixed effect}  & Random effect\\
 \cline{2-4}
 Experiment & \multicolumn{2}{l}{Estimate}  &Std. error & Variance\\
  \hline
  1 & 0&39 & 0.0094 & 0.0080 \\ 
  2 & 1&21 &  0.0197 &  0.0443 \\ 
  3 & 0&59 (Intercept)  &   0.1668 & 1.9917\\ 
     & 0&21 (Slope)    &  0.0511     &  0.0245\\ 
     &-0&78 (correlation) & & \\
   \hline
\end{tabular}
\end{center}
\label{tbl:model_par}
\end{table}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect.pdf}}
       \caption{Power in comparison to effect for the three experiments. Points indicate subject responses, with size indicating count. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. The loess curve (blue) estimates the observed power, and grey bands show simultaneous bootstrap confidence band.  Conventional test power is drawn as a dashed line. For experiment 3, conventional power is based on the slopes of the linear part of the non-contaminated data. Power for contaminated data is shown by cross mark.}
       \label{fig:power_loess_effect}
\end{figure}

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
 The estimated subject-specific power curves are shown in Figure \ref{fig:power_mixed_subject} (light grey lines). The blue curve is the overall average power estimated from the model, and for comparison, the grey dashed lines are the power curves for the conventional test. Subject-specific power is quite different for three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Investigating how does the strength of structure in the null plots affects the choice of the actual data plot. Difference between the data plot's $p$-value and the  smallest $p$ value of a null plot vs proportion correct. Negative values indicate the $p$-value of the actual data plot are smaller than those of all of the null plots. Close to zero shows a wide range in the proportion correct, suggesting that when at least one null plot has structure almost as strong as the actual data plot, subjects had a difficult time in making their choice.}
       \label{fig:pval_difference}
\end{figure}

@@

Experimental subjects in the simulation study came from various level of education, both genders and many different ages. Many subjects had little background in statistics, and were not necessarily aware of statistical plots. Yet the results were good. We expect that visual inference could a useful tool for high school students to learn about randomness as a primer to statistical thinking. A major advantage of visual statistical inference is that it is intuitive, as illustrated by the performance of subjects with limited education, and it is may make statistics fun to learn.

>>>>

\begin{table*}[hbtp]
\caption{Comparison of visual inference with conventional inference.}
\centering 
\begin{tabular}{llll} 
\hline \hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\scalebox{.9}{
\begin{tabular}{rrrrrrrrrrrrrrrrrrr}
$K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$  & $x$ & $p$-value && $K$  & $x$ & $p$-value\\ \cline{1-3} \cline{5-7}\cline{9-11}\cline{13-15}\cline{17-19}
1 &  1 & 0.0500 && 2 &  1 & 0.0975 && 3 & 1 & 0.1426 && 4 & 1 & 0.1855 && 5 & 1 & 0.2262 \\%\cline{1-3}
&&&& 2 &  2 & 0.0025 && 3 & 2 & 0.0073 && 4 & 2 & 0.0140 && 5 & 2 & 0.0226 \\
&&&& &&&& 3 & 3 & 0.0001 && 4 & 3 & 0.0005 && 5 & 3 & 0.0012 \\%\cline{1-6}
&&&&  &    &        &   &   &        &&& 4 & 4 & $< 0.0001$ && 5 & 4 & $< 0.0001$ 
\end{tabular}
}
\end{center}
\label{pvalue}
\end{table}

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of size $\alpha$, is defined to be 
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot in the lineup, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
\end{dfn}

@@

\hh{If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p_i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition, see \citet{butler93} for details. Estimating $p_i$ will be one of the main concerns for the remainder of the paper.} One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

@@

Under the null hypothesis $X \sim Binom_{K, 1/m}$, leading to a type I error of:
\[
\text{Power}_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
where $x_\alpha$ is defined as in (\ref{dfn:test}), as  $P(X \ge x_{\alpha}) \le \alpha$.

@@

%\green{Notation changes $i$ to $j$, and $p$ to $\pi$. Can someone please decide which to use, and make the appropriate changes? How does the ``Let $X_{j\ell} \sim B_{1, \pi_{j\ell}}$...'' below relate to the ``let $X_i$ be a binary random variable...'' above? It looks like this is a repeat. One needs to go once notation is clarified.}

%\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate actual data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

@@

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for each of the simulation experiments.} % title name of the table
\centering
\scalebox{.9}{
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

% Entering 2rd row
&  5 & 0, 1, 2, 3, 5  & 0.1, 0.4, 0.7, 1, 1.5&\\[-1ex]
\raisebox{1.5ex}{300} & 12
& 1, 3, 5, 7, 10  & 0, 0.8, 1.75, 2.3, 3.5 &\\[1ex]
% [1ex] adds vertical space
\hline
% inserts single-line
\end{tabular}
}
\label{tbl:experiment_params}
\end{table}

@@

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }%\blue{Notice that we do not have data on one lineup (so we have only 29 lineups) for experiment 3 due to the screening process.}}
\begin{center}
\begin{tabular}{ccrrcc}
   \hline \hline
 Experiment & Subject & Male & Female & Responses &Lineup\\ 
    \hline
1 & 239 & 121 & 107 & 2249 &  60 \\ 
  2 & 351 & 185 & 164 & 3636 &  70 \\ 
  3 & 155 & 103 &  52 & 1511 &  29 \\  
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

\begin{table}[hbtp]
\caption{Parameter estimates of model in Equation \ref{eqn:mixed}. Estimates are highly significant with $p$-value $<$  0.0001 for all three experiment data.}
\begin{center}
\begin{tabular}{cr@{.}lcc}
  \hline \hline
 &  \multicolumn{3}{c} {Fixed effect}  & Random effect\\
 \cline{2-4}
 Experiment & \multicolumn{2}{l}{Estimate}  &Std. error & Variance\\
  \hline
  1 & 0&39 & 0.0094 & 0.0080 \\ 
  2 & 1&21 &  0.0197 &  0.0443 \\ 
  3 & 0&59 (Intercept)  &   0.1668 & 1.9917\\ 
     & 0&21 (Slope)    &  0.0511     &  0.0245\\ 
     &-0&78 (correlation) & & \\
   \hline
\end{tabular}
\end{center}
\label{tbl:model_par}
\end{table}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect.pdf}}
       \caption{Power in comparison to effect for the three experiments. Points indicate subject responses, with size indicating count. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. \blue{The loess curve (blue) estimates the observed power}, and grey bands show simultaneous bootstrap confidence band.  Conventional test power is drawn as a dashed line. For experiment 3, conventional power is based on the slopes of the linear part of the non-contaminated data. Power for contaminated data is shown by cross mark.}
       \label{fig:power_loess_effect}
\end{figure}

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
 The estimated subject-specific power curves are shown in \blue{Figure \ref{fig:power_mixed_subject} (light grey lines). The blue curve is the overall} average power estimated from the model, and for comparison, the grey dashed lines are the power curves for the conventional test. Subject-specific power is quite different for three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Investigating how the strength of structure in the null plots affects the choice of the actual data plot. Difference between the data plot's $p$-value and the  smallest $p$ value of a null plot vs proportion correct. Negative values indicate the $p$-value of the actual data plot are smaller than those of all of the null plots. Close to zero shows a wide range in the proportion correct, suggesting that when at least one null plot has structure almost as strong as the actual data plot, subjects had a difficult time in making their choice.}
       \label{fig:pval_difference}
\end{figure}

@@

Experimental subjects in the simulation study came from various level of education, both genders and many different ages. Many subjects had little background in statistics, and were not necessarily aware of statistical plots. Yet the results were good. We expect that visual inference could be a useful tool for high school students to learn about randomness as a primer to statistical thinking. A major advantage of visual statistical inference is that it is intuitive, as illustrated by the performance of subjects with limited education, and it is may make statistics fun to learn.
